Revisiting~the~Classics: Online~RL~in~the~Programmable~Dataplane

### 2021-06-08 paper/paper.tex
Now that's what I call progress.

@@

\begin{figure*}
	\centering
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[keepaspectratio, width=0.85\linewidth]{figures/ind}
		\caption{\Indfw{} (offline throughput-optimal). \emph{Workers} independently pull commands from (and push state-action pairs to) the environment, locking access to the policy when updates must occur.\label{fig:single-and-parallel:single}}
	\end{subfigure}
	\hspace{0.04\linewidth}
	\begin{subfigure}{0.45\linewidth}
		\centering
		\includegraphics[keepaspectratio, width=0.85\linewidth]{figures/coop}
		\caption{\Coopfw{} (online-optimal). A single \emph{controller} delegates RL computation and updates to many \emph{minion} threads, who operate on independent subtasks.\label{fig:single-and-parallel:parallel}}
	\end{subfigure}
	\caption{\approachshort{}'s compute strategies scale to fit device capacity according to either latency or throughput needs.\label{fig:single-and-parallel}}
\end{figure*}

@@

\subsection{Results and Discussion}\label{sec:results}
\fakepara{Raw inference and learning performance}
\Cref{tab:lats} shows how \approachshort{} compares in latency with a numpy-based implementation of an RL policy\footnote{For brevity, we omit numpy-based integer results---summarised against a floating point implementation, median state-action latencies are \SI{14.6}{\percent} worse, with \SI{7.9}{\percent} longer update times.}.
We show that \Coopfw{} achieves sub-\SI{35}{\micro\second} median latency, with \nth{99} and \num{99.99}\nthscript{th} percentile latencies less than \SI{1}{\micro\second} worse using 4 MEs of the NFP-6480.
This corresponds to \SIlist{15.0;21.0}{$\times$} speedups over a \emph{Collector} host.
Importantly, \Indfw{} achieves lower median state-action latencies (\SI{2.79}{$\times$}) \emph{and} update times (\SI{2.63}{$\times$}) than a dedicated \emph{Collector} host \emph{while requiring only a single core or dedicated functional unit}.

@@

\Cref{tab:tputs} compares \approachshort{}'s throughput against host-based execution.
We choose the worker count on host machines that maximises throughput.
This equalled the amount of physical cores on each device, while moving beyond this (even below the number of hyper-threads) would hamper tail latencies by an order of magnitude.
To make the comparison fair in the context of many-core CPU environments, we include per-core throughput.
\Indfw{} achieves \SI{2.82}{$\times$} higher offline throughput than commodity \emph{Collector} hardware in spite of the NFP-6480 having a considerably slower clock speed (\SI{0.29}{$\times$}).
When compounded with the abundance of such weaker chips, in-NIC RL is able to deliver much higher throughput.
%As anticipated, the \Coopfw{} strategy is key in achieving serviceable throughput in an online learning agent, \SI{9.9}{$\times$} that of a dedicated collector machine, due to the locking requirements mandating coordinated division of work.
As anticipated, the \Coopfw{} strategy is key in achieving serviceable throughput in an online learning agent, \SI{9.9}{$\times$} that of a dedicated collector machine, as the write lock around policy updates creates a bottleneck.

By limiting the available workers in software, we show how \Coopfw{}'s policy update time (thus  online throughput---\cref{fig:vary-core}) and state-action latency (\cref{fig:vary-core-latency}) scale with available cores.
While the compute times and latency of \Coopfw{} always outperform the host-based floating point implementations, we observe that there are two distinct crossover points which must be met to overcome our own \Indfw{}; \num{8} workers for online throughput, and \num{3} workers for state-action latency.
Some artefacts of our environment and design choices are visible, such as the addition of new physical cores (MEs) being more significant than contexts, and the presence of some schedule bottlenecks.
Most importantly however, \approachshort{}-\Coopfw{}'s resource demand is tunable at compile time to meet the online training rate and/or action latency required by a task/environment.

\begin{figure}
	\resizebox{0.9\linewidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core}
	}
	\caption{
		\Coopfw{}'s online learning performance improves as additional cores are made available, on max size tasks (\num{129} work items). This requires \num{8} workers to offer greater online throughput than single-threaded in-NIC RL. Sharper performance increases occur when a new physical core is added (\numrange{7}{8}) or the scheduler works around a bottleneck (\numrange{13}{14}).\label{fig:vary-core}}
\end{figure}

\begin{figure}
	\resizebox{0.9\linewidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core-latency}
	}
	\caption{\Coopfw{}'s action latencies improve with additional cores. This requires \num{3} workers (4 total contexts) to improve upon the state-action latency of single-threaded in-NIC RL.\label{fig:vary-core-latency}}
\end{figure}

\Cref{fig:vary-work,fig:vary-work-latency} show how policy complexity affects update costs and state-action latency respectively, scaling from a bias tile up to the full DDoS policy size\footnote{Input vectors here all have 20 elements regardless of the policy design.}.
\Coopfw{} always produces an action in less time than \Indfw{}, but requires at least one state-based tile to excel in online learning.
%We note that this is a trivial case, in that the use of \emph{only} a bias tile negates the need for any input state (similar to a multi-arm bandit problem).
We note that this is a trivial case, as using \emph{only} a bias tile returns a single preference list regardless of input state.

\begin{figure}
	\resizebox{0.9\linewidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-work}
	}
	\caption{\Coopfw{} fully processes updates faster than \Indfw{}---thus has higher online performance---on almost all policy sizes. Lower bit depths are more effective on simpler policies.\label{fig:vary-work}}
\end{figure}

We found that a key aspect of in-NIC execution is that it allows far tighter bounds on tail latency compared to host offloading.
Examining the state-action latencies in \cref{tab:lats}, we see that \num{99.99}\nthscript{th} percentile latencies exceed the median by \SIlist{0.58;0.66}{\percent} for \Indfw{} and \Coopfw{}, respectively.
Similar results were observed for other bit depths.
By contrast, host-based tail latencies are at least \SI{40.53}{\percent} greater even when parallel workers are at or below the physical core count.
We show the cumulative distributions of these in detail (\cref{fig:lat-cumul}), noting how just one additional CPU-intensive task---potentially automated system updates, scans, or another vNF/traffic processing task---impacts tail latencies further (\emph{Float(Over)}).

\begin{figure}
	\resizebox{0.9\linewidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-work-latency}
	}
	\caption{State-action latency scales with additional work in a similar manner to overall processing time; though \SI{32}{\bit} firmwares become more effective sooner (at \num{3} work items).\label{fig:vary-work-latency}}
\end{figure}

\begin{figure}
	\centering
	\begin{subfigure}{0.45\linewidth}
		\resizebox{1.0\linewidth}{!}{
			\includegraphics{../plots/build/rl-perf-tester/latency-cumul}
		}
		\caption{\approachshort's \Coopfw{} design achieves consistent, tight latency bounds.}
	\end{subfigure}
	\begin{subfigure}{0.45\linewidth}
		\resizebox{1.0\linewidth}{!}{
			\includegraphics{../plots/build/rl-perf-tester/latency-cumul-broad}
		}
		\caption{Tail latencies suffer in host execution---particularly when oversubscribed.}
	\end{subfigure}
	\caption{Cumulative state-action latency plots for \approachshort{} and host-based execution.\label{fig:lat-cumul}}
\end{figure}

A noteworthy trend is that the \SIlist{8;16}{\bit} versions of \approachshort{} consistently underperform compared to \SI{32}{\bit} designs, with the exception of smaller workloads (seen in the zoomed portions of \cref{fig:vary-work,fig:vary-work-latency}).
This held even though our implementation is optimised to read and write policy tile data in batches (achieving fewer I/O operations as required).
This is due to the native register width on the NFP being \SI{32}{\bit}, and so the compiler must emit extra instructions around arithmetic operations to correctly load and update values.
This matches \SI{32}{\bit} becoming dominant in more complex workloads: higher dimension tilings require more arithmetic operations to compute the hit tile index.
Most of the I/O comes after this step, causing ALU use to dominate.
This also explains why \SI{32}{\bit} becomes the best choice at different policy complexities for online (\cref{fig:vary-work}, 10 dims) and offline (\cref{fig:vary-work-latency}, 3 dims) \Coopfw{} agents, where hashtable accesses and a \texttt{memcpy} of the state vector fall into the serial portion of the algorithm in the online case.
As a result, these bit depths give a \SIrange{2}{4}{$\times$} reduction in memory needed to store a policy (enabling more fine-grained or more complex policies), in exchange for slightly worse latency and throughput.
We investigated bit-stuffing several such values into a single word for our atomic writeback mechanism (as the platform offers both \SIlist{32;64
}{\bit} atomic addition).
This is analogous to SIMD---through clever use of padding bits---but we found that manipulating tiles into the correct format added \SI{10}{\percent} extra overhead.

\fakepara{Work allocation}
\Cref{fig:work-alloc-32} shows that our heuristic first-fit allocator (\emph{Balanced}) is key in achieving consistent sub-\SIlist{35;65}{\micro\second} latencies/processing times, respectively.
The trend is repeated for all bit depths, noting the latency differences described above.
The constant difference between \emph{Action} and \emph{Update Prep} was observed to scale linearly with bit depth, matching storage and lookup work in the serial portion between parallel jobs (plots omitted).
The severe underperformance of the \emph{Na\"{i}ve} allocator confirms that work item complexity is correlated with its ID, as batching work in contiguous chunks leads to some workers receiving exclusively high-dimensionality tile sets.
The minor gap in performance lower bounds between the \emph{Random} and \emph{Balanced} allocators suggests that further optimisations can be made in this way.
We expect that closing or exceeding this gap may require more complex modelling of NFP's hardware thread interactions, which lies far beyond the scope of in-NIC scheduling.
Some additional complexity may be tolerated, subject to code store limits---scheduling runs exactly once per configuration change, so does not impact per-action code.

\begin{figure}
	\resizebox{0.9\linewidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/work-strat-32bit}
	}
	\caption{Action/update compute times in a \SI{32}{\bit} \Coopfw{} agent under different work schedulers.\label{fig:work-alloc-32}}
\end{figure}

\begin{figure}
	\resizebox{0.9\linewidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core-tput-on-per}
	}
	\caption{Throughput per added worker in a \Coopfw{} agent.\label{fig:tput-per-core}}
\end{figure}

@@

\fakepara{End-to-end RL latency}
%?? PCIe RTT is 10us (neugebauer, BNN NFP paper), NFP RTT is 18us (own measurements). EMEM Ring one-way is 120ns.
%?? Cziva~\parencite[pp.113]{DBLP:phd/ethos/Cziva18} discusses vNF times.
%?? reference the inference times brought up in Taurus?
Referring to \cref{fig:state-slip}, we take $t_2$ from \cref{tab:lats} for host and in-NIC processing times, and substitute $t_1+t_3$ for the state packet round-trip time to the decision site:
\begin{description}
	\item[In-NIC.] As described in \cref{sec:agent-environment-communication}, EMEM rings have a median one way delay cross-island of \SI{140}{\nano\second}, giving a \emph{median \SI{34.63}{\micro\second} end-to-end inference latency}.
	\item[Dedicated Collector.] Offloads hosted in this manner will employ DPDK to maximise performance, giving one-way PCIe delays of \SIrange{0.9}{2.3}{\micro\second} for network packets.
	A UDP packet carrying \num{20} elements of state in \approachshort{} is \SI{128}{\byte}, so costs \SI{1}{\micro\second} to forward, and the reply state-action pair invokes a slightly higher cost.
	\emph{This gives an end-to-end inference latency of \SI{517.9}{\micro\second}}.
	\item[vNF Offload.] \Textcite{DBLP:journals/cm/CzivaP17} show that more lightweight vNF frameworks like GNF~\parencite{DBLP:journals/cm/CzivaP17} and ClickOS~\parencite{DBLP:conf/nsdi/MartinsAROHBH14} add \SIrange{45}{55}{\micro\second} \emph{additional} latency above PCIe costs.
	\emph{This gives an end-to-end inference latency of \SI{572.9}{\micro\second}}.
\end{description}
Using these estimates, in-NIC classical RL inference offers \SIlist{14.96;16.54}{$\times$} speedups in latency over collector and vNF deployments respectively (assuming no steering cost in either case).
We also contrast these against deep RL policies on network tasks, which can take \SI{3}{\milli\second} to compute~\parencite{DBLP:journals/corr/abs-1910-04054}---2 orders of magnitude above \approachshort{} with identically sized (20-dim) input vectors.
%In the name of fairness, we assume that rule installation uses same mechanism we recommend, but show how bad it can get? I.e. with rule installation costs etc.

\fakepara{Co-existence with the dataplane}
We were able to saturate the \SI{40}{\giga\bit\per\second} link for packet sizes at or above \SI{256}{\byte}.
%For frame sizes of \SIlist{64;128;256}{\byte} input traffic rates were \SIlist{17.4;32.9;37.0}{\giga\bit\per\second} respectively (\SIlist[per-symbol=p,sticky-per=true]{33.9;32.2;18.1}{\mega\packet\per\second}).
For frame sizes of \SIlist{64;128}{\byte} input traffic rates were \SIlist{17.4;32.9}{\giga\bit\per\second} respectively (\SIlist[per-symbol=p,sticky-per=true]{33.9;32.2}{\mega\packet\per\second}).
Passing this traffic over the loaded NFP device running \approachshort, no packet losses were incurred at any rate of RL actions.

We show the effect of RL workloads on the round-trip latencies of cross traffic through \cref{fig:dataplane-coop}.
As observed latency measures do not obey a normal distribution (particularly \SIlist{256;1518}{\byte}, which are bimodal), we employ a one-tailed \emph{Mann-Whitney U test} to mark statistically significant population increases in latency ($p < 0.05$) with a ``+''.
In general, we observe that statistically significant latency increases concentrate around smaller packet sizes.
All (bar one) of these affected \nth{99} percentile latencies by under \SI{0.38}{\percent} (at most \SI{78}{\nano\second}).
This slight degradation can be explained by increased pressure on the NFP's \emph{Command Push-Pull} (CPP) bus, which is responsible for handling cross-island accesses to memory (particularly IMEM/EMEM) and other resources.
\approachshort{} places load on the CPP bus through use of its \inring{}/\outring{} rings for external communication and through last-tier policy accesses.
This also explains the sensitivity of \SI{256}{\byte} packets to \approachshort{}---Netronome's P4 toolchain segments packets, storing metadata (e.g., MAC prepend data) and the first bytes of a packet in a $\SI{256}{\byte}$ CTM block and parking their payloads in EMEM.
\SI{256}{\byte} packets overshoot this due to metadata, causing small I/O accesses at a high rate for packets sized around this cutoff point.
Regardless, this effect is small when observed.

@@

\begin{figure}
	\centering
	\begin{subfigure}{0.45\linewidth}
		\resizebox{1.0\linewidth}{!}{
%			\includegraphics{../plots/build/stress/heat-latency-median}
			\includegraphics{../plots/build/stress/heat-latency-two-9-rel}
%			\includegraphics{../plots/build/stress/heat-latency-two-9-perc}
		}
		\caption{Deviations in \nth{99} percentile cross-traffic RTTs.\label{fig:dataplane-heat}}
	\end{subfigure}
	\hspace{0.05\linewidth}
		\begin{subfigure}{0.45\linewidth}
		\resizebox{1.0\linewidth}{!}{
			\includegraphics{../plots/build/stress/histo-128B-0-3-trim}
		}
		\caption{Distribution of RTTs for \SI{128}{\byte} packets for baseline and \num{3000} RL actions/s.\label{fig:dataplane-example}}
	\end{subfigure}
	\caption{Effects on tail latency of cross-traffic caused by different loads of off-path RL compute. Statistically significant increases in population latency are concentrated on smaller packet sizes, and are typically sub-\SI{78}{\nano\second}.\label{fig:dataplane-coop}}
\end{figure}

\fakepara{Resource requirements}
\Cref{tab:resources} demonstrates how \approachshort{} consumes shared and local memory as it scales to fit a device's compute resources, compared with a simple P4 forwarding application.
As one program is installed per ME, these results represent the minimum and maximum resource use we can command on a single island (i.e., without replacing P4 workers).
We observe negligible costs on shared bulk EMEM storage ($\sim$\SI{4}{\mebi\byte}), including storage space for a hash-table of \num{4096} state-action pairs and \num{16} reward values.
The most significant costs arise due to policy data (\SI{405}{\kibi\byte} shared IMEM, \SI{90}{\kibi\byte} local CTM, \SI{15}{\kibi\byte} local CLS), which can be halved or quartered using \SIlist{16;8}{\bit} quantisation and remain constant regardless of compute unit usage.
This causes a high upfront cost on per-island resources (CLS/CTM)---\approachshort{} leaves resources for other off-path dataplane applications, but is fairest from 3 cores onwards.

@@

\fakepara{Why in-NIC?}
The above approach relies upon co-hosting traffic measurement and analysis alongside OpenFlow-compatible switches at the edge nodes of an AS.
However, packet mirroring, offloading RL computation to a host (potentially over layers of virtualisation), and CPU performance are all sources of additional, consistent state-action latency.
Both traffic statistics collection \emph{and} data-driven learning must be executed on such hosts/network functions.
Unless running these stages in a dedicated pipeline (adding further processing latency), resource contention between these processes will further impact tail latencies.
Naturally, this requires high-performance hardware to be stacked at network egress points, potentially beyond reasonable space, power, or ventilation constraints.
The solution to implement and improve upon this work using \approachshort{} is to place its RL agents on SmartNICs at AS edge nodes---a bump-in-the-wire deployment.

>>>>

\setlength{\abovecaptionskip}{5pt}
\setlength{\belowcaptionskip}{0pt}

\setlength{\textfloatsep}{6.0pt plus 2.0pt minus 4.0pt}
\setlength{\floatsep}{6.0pt plus 1.0pt minus 2.0pt}

\setlength{\dbltextfloatsep}{6.0pt plus 2.0pt minus 4.0pt}
\setlength{\dblfloatsep}{6.0pt plus 1.0pt minus 2.0pt}

@@

\begin{figure*}
	\centering
	\begin{subfigure}{0.47\linewidth}
		\centering
		\includegraphics[keepaspectratio, width=0.78\linewidth]{figures/ind}
		\caption{\Indfw{} (offline throughput-optimal). \emph{Workers} independently pull commands from (and push actions to) the environment, locking policy access for updates.\label{fig:single-and-parallel:single}}
	\end{subfigure}
	\hspace{0.04\linewidth}
	\begin{subfigure}{0.47\linewidth}
		\centering
		\includegraphics[keepaspectratio, width=0.8\linewidth]{figures/coop}
		\caption{\Coopfw{} (online-optimal). A single \emph{controller} delegates RL computation and updates to many \emph{minion} threads, who operate on independent subtasks.\label{fig:single-and-parallel:parallel}}
	\end{subfigure}
	\caption{\approachshort{}'s compute strategies scale to fit device capacity according to either latency or throughput needs.\label{fig:single-and-parallel}}
\end{figure*}

@@

\subsection{Results and Discussion}\label{sec:results}
\fakepara{Raw inference and learning performance}
\Cref{tab:lats} shows how \approachshort{} compares in latency with a numpy-based implementation of an RL policy\footnote{For brevity, we omit numpy-based integer results---summarised against a floating point implementation, median state-action latencies are \SI{14.6}{\percent} worse, with \SI{7.9}{\percent} longer update times.}.
We show that \Coopfw{} achieves sub-\SI{35}{\micro\second} median latency, with \nth{99} and \num{99.99}\nthscript{th} percentile latencies less than \SI{1}{\micro\second} worse using 4 MEs of the NFP-6480.
This corresponds to \SIlist{15;21}{$\times$} speedups over a \emph{Collector} host.
Importantly, \Indfw{} achieves lower median state-action latencies (\SI{2.79}{$\times$}) \emph{and} update times (\SI{2.63}{$\times$}) than a dedicated \emph{Collector} host \emph{while requiring only a single core or dedicated functional unit}.

\newlength{\resultplotwidth}
\setlength{\resultplotwidth}{0.85\linewidth}

@@

\Cref{tab:tputs} compares \approachshort{}'s throughput against host-based execution.
We chose the worker count on host machines that maximised their throughput.
This equalled the amount of physical cores on each device---moving beyond this (even below the number of hyper-threads) would hamper tail latencies by an order of magnitude.
To make the comparison fair in the context of many-core CPU environments, we include per-core throughput.
\Indfw{} achieves \SI{2.82}{$\times$} higher offline throughput than commodity \emph{Collector} hardware in spite of the NFP-6480 having a considerably slower clock speed (\SI{0.29}{$\times$}).
When compounded with the abundance of such weaker chips, in-NIC RL is able to deliver much higher throughput.
%As anticipated, the \Coopfw{} strategy is key in achieving serviceable throughput in an online learning agent, \SI{9.9}{$\times$} that of a dedicated collector machine, due to the locking requirements mandating coordinated division of work.
As anticipated, the \Coopfw{} strategy is key in achieving serviceable throughput in an online learning agent, \SI{9.9}{$\times$} that of a dedicated collector machine, as the write lock around policy updates creates a bottleneck.

By limiting the available workers in software, we show how \Coopfw{}'s policy update time (thus  online throughput---\cref{fig:vary-core}) and state-action latency (\cref{fig:vary-core-latency}) scale with available cores.
While the compute times and latency of \Coopfw{} always outperform the host-based floating point implementations, we observe that there are two distinct crossover points which must be met to overcome our own \Indfw{}; \num{8} workers for online throughput, and \num{3} workers for state-action latency.
Some artefacts of our environment and design choices are visible, such as the addition of new physical cores being more significant than contexts, and the presence of some schedule bottlenecks.
Most importantly however, \approachshort{}-\Coopfw{}'s resource demand is tunable at compile time to meet the online training rate and/or action latency required by a task/environment.

\begin{figure}
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core}
	}
	\caption{
		\Coopfw{}'s online learning performance improves as additional cores are made available, on max size tasks (\num{129} work items). This requires \num{8} workers to offer greater online throughput than single-threaded in-NIC RL. Sharper performance increases occur when a new physical core is added (\numrange{7}{8}) or the scheduler works around a bottleneck (\numrange{13}{14}).\label{fig:vary-core}}
\end{figure}

\begin{figure}
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core-latency}
	}
	\caption{\Coopfw{}'s action latencies improve with additional cores. This requires \num{3} workers (4 total contexts) to improve upon the state-action latency of single-threaded in-NIC RL.\label{fig:vary-core-latency}}
\end{figure}

\Cref{fig:vary-work,fig:vary-work-latency} show how policy complexity affects update cost and state-action latency respectively, scaling from a bias tile up to the full DDoS policy size\footnote{Input vectors here all have 20 elements regardless of the policy design.}.
\Coopfw{} always produces an action in less time than \Indfw{}, but requires at least one state-based tile to excel in online learning.
%We note that this is a trivial case, in that the use of \emph{only} a bias tile negates the need for any input state (similar to a multi-arm bandit problem).
We note that this is a trivial case, as using \emph{only} a bias tile returns a single preference list regardless of input state.

\begin{figure}
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-work}
	}
	\caption{\Coopfw{} fully processes updates faster than \Indfw{}---thus has higher online performance---on almost all policy sizes. Lower bit depths are more effective on simpler policies.\label{fig:vary-work}}
\end{figure}

We found that a key aspect of in-NIC execution is that it allows far tighter bounds on tail latency compared to host offloading.
Examining the state-action latencies in \cref{tab:lats}, we see that \num{99.99}\nthscript{th} percentile latencies exceed the median by \SIlist{0.58;0.66}{\percent} for \Indfw{} and \Coopfw{}, respectively.
Similar results were observed for other bit depths.
By contrast, host-based tail latencies are at least \SI{40.53}{\percent} greater even when the parallel worker count is at or below the physical core count.
We show the cumulative distributions of these in detail (\cref{fig:lat-cumul}), noting how just one additional CPU-intensive task---potentially automated system updates, scans, or another vNF/traffic processing task---impacts tail latencies further (\emph{Float(Over)}).

\begin{figure}
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-work-latency}
	}
	\caption{State-action latency scales with additional work in a similar manner to overall processing time; though \SI{32}{\bit} firmwares become more effective sooner (at \num{3} work items).\label{fig:vary-work-latency}}
\end{figure}

\begin{figure}
	\centering
	\begin{subfigure}{0.45\linewidth}
		\resizebox{1.0\linewidth}{!}{
			\includegraphics{../plots/build/rl-perf-tester/latency-cumul}
		}
		\caption{\approachshort's \Coopfw{} design achieves consistent, tight latency bounds.}
	\end{subfigure}
	\hspace{0.05\linewidth}
	\begin{subfigure}{0.45\linewidth}
		\resizebox{1.0\linewidth}{!}{
			\includegraphics{../plots/build/rl-perf-tester/latency-cumul-broad}
		}
		\caption{Tail latencies suffer in hosts---particularly when oversubscribed.}
	\end{subfigure}
	\caption{Cumulative state-action latency plots for \approachshort{} and host-based execution.\label{fig:lat-cumul}}
\end{figure}

A noteworthy trend is that the \SIlist{8;16}{\bit} versions of \approachshort{} consistently underperform compared to \SI{32}{\bit} designs, with the exception of smaller workloads (seen in the zoomed portions of \cref{fig:vary-work,fig:vary-work-latency}).
This occurs even though our implementation is optimised to read and write policy tile data in batches (achieving fewer I/O operations as required).
We see this because the native register width on the NFP is \SI{32}{\bit}, and so the compiler must emit extra instructions around arithmetic operations to correctly load and update values.
This matches \SI{32}{\bit} becoming dominant in complex workloads: higher dimension tilings require more arithmetic operations to compute.
Most of the I/O comes \emph{after} this step, causing ALU use to dominate.
This also explains why \SI{32}{\bit} becomes the best choice at different policy complexities for online (\cref{fig:vary-work}, 10 dims) and offline (\cref{fig:vary-work-latency}, 3 dims) agents, where hashtable accesses and a \texttt{memcpy} of the state vector fall into the serial portion of the online algorithm.
Smaller bit depths still give a \SIrange{2}{4}{$\times$} saving in memory for policy storage (enabling more fine-grained or more complex policies), in exchange for slightly worse latency and throughput.
To overcome this, we investigated bit-stuffing several values into a single word for our atomic writeback mechanism (as the platform offers both \SIlist{32;64
}{\bit} atomic addition).
This is analogous to SIMD through clever use of padding bits, but we found that manipulating tiles into the correct format added \SI{10}{\percent} extra overhead.

\fakepara{Work allocation}
\Cref{fig:work-alloc-32} shows that our heuristic allocator (\emph{Balanced}) is key in achieving consistent sub-\SIlist{35;65}{\micro\second} latencies/update times, respectively.
The trend is repeated for all bit depths.
The constant difference between \emph{Action} and \emph{Update Prep} was observed to scale linearly with bit depth, matching storage and lookup work in the serial portion of \emph{ParSa} (plots omitted).
The severe underperformance of the \emph{Na\"{i}ve} allocator confirms that work item complexity is correlated with its index, as batching work in contiguous chunks leads to some workers receiving exclusively high-dimensionality tilings.
The minor gap in lower bound performance between the \emph{Random} and \emph{Balanced} allocators suggests that further optimisations can be made.
We expect that closing or exceeding this gap may require more complex modelling of hardware thread interactions, which lies far beyond the scope of in-NIC scheduling.
Some additional complexity may be tolerated, subject to code store limits---scheduling runs exactly once per configuration change, so does not impact per-action code.

\begin{figure}
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/work-strat-32bit}
	}
	\caption{Action/update compute times in a \SI{32}{\bit} \Coopfw{} agent under different work schedulers.\label{fig:work-alloc-32}}
\end{figure}

\begin{figure}
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core-tput-on-per}
	}
	\caption{Throughput per added worker in a \Coopfw{} agent.\label{fig:tput-per-core}}
\end{figure}

@@

\fakepara{End-to-end RL latency}
%?? PCIe RTT is 10us (neugebauer, BNN NFP paper), NFP RTT is 18us (own measurements). EMEM Ring one-way is 120ns.
%?? Cziva~\parencite[pp.113]{DBLP:phd/ethos/Cziva18} discusses vNF times.
%?? reference the inference times brought up in Taurus?
Referring to \cref{fig:state-slip}, we take $t_2$ from \cref{tab:lats} for host and in-NIC processing times, and substitute $t_1+t_3$ for the state packet round-trip time to the decision site:
\begin{description}
	\item[In-NIC.] As described in \cref{sec:agent-environment-communication}, EMEM rings have a median one way delay cross-island of \SI{140}{\nano\second}, giving a \emph{median \SI{34.63}{\micro\second} end-to-end inference latency}.
	\item[Dedicated Collector.] Offloads hosted in this manner will employ DPDK to maximise performance, giving one-way PCIe delays of \SIrange{0.9}{2.3}{\micro\second} for network packets.
	A UDP packet carrying \num{20} elements of state in \approachshort{} is \SI{128}{\byte}, so costs \SI{1}{\micro\second} to forward, and the reply state-action pair invokes a slightly higher cost.
	\emph{This gives an end-to-end inference latency of \SI{517.9}{\micro\second}}.
	\item[vNF Offload.] \Textcite{DBLP:journals/cm/CzivaP17} show that more lightweight vNF frameworks like GNF~\parencite{DBLP:journals/cm/CzivaP17} and ClickOS~\parencite{DBLP:conf/nsdi/MartinsAROHBH14} add \SIrange{45}{55}{\micro\second} \emph{additional} RTT latency above PCIe costs.
	\emph{This gives an end-to-end inference latency of \SI{572.9}{\micro\second}}.
\end{description}
Using these estimates, in-NIC classical RL inference offers \SIlist{14.96;16.54}{$\times$} speedups in latency over collector and vNF deployments respectively (assuming no steering cost in either case).
We also contrast these against deep RL policies on network tasks, which can take \SI{3}{\milli\second} to compute~\parencite{DBLP:journals/corr/abs-1910-04054}---2 orders of magnitude above \approachshort{} with identically sized (20-dim) input vectors.
%In the name of fairness, we assume that rule installation uses same mechanism we recommend, but show how bad it can get? I.e. with rule installation costs etc.

\fakepara{Co-existence with the dataplane}
Our setup saturated the \SI{40}{\giga\bit\per\second} link for packet sizes $\ge$\SI{256}{\byte}.
%For frame sizes of \SIlist{64;128;256}{\byte} input traffic rates were \SIlist{17.4;32.9;37.0}{\giga\bit\per\second} respectively (\SIlist[per-symbol=p,sticky-per=true]{33.9;32.2;18.1}{\mega\packet\per\second}).
For frame sizes of \SIlist{64;128}{\byte}, input traffic rates were \SIlist{17.4;32.9}{\giga\bit\per\second} respectively (\SIlist[per-symbol=p,sticky-per=true]{33.9;32.2}{\mega\packet\per\second}).
Passing this traffic over the NFP device running \approachshort, no packet losses were incurred at any rate of RL actions.

We show the effect of RL workloads on the round-trip latencies of cross traffic through \cref{fig:dataplane-coop}.
As observed latencies do not obey a normal distribution (particularly \SIlist{256;1518}{\byte}, which are bimodal), we employ a one-tailed \emph{Mann-Whitney U test} to mark statistically significant population increases in latency ($p < 0.05$) with a ``+''.
In general, statistically significant latency increases concentrate around smaller packet sizes.
All (bar one) of these affected \nth{99} percentile latencies by under \SI{0.38}{\percent} (at most \SI{78}{\nano\second}).
This slight degradation can be explained by increased pressure on the NFP's \emph{Command Push-Pull} (CPP) bus, which is responsible for handling cross-island accesses to memory (particularly IMEM/EMEM) and other resources.
\approachshort{} places load on the CPP bus through its \inring{}/\outring{} EMEM rings and last-tier policy accesses.
This also explains the sensitivity of \SI{256}{\byte} packets to \approachshort{}---the NFP P4 toolchain segments packets, storing metadata (e.g., MAC prepend) and the first bytes of a packet in a \SI{256}{\byte} CTM block and parking their payloads in EMEM.
\SI{256}{\byte} packets overshoot this due to metadata, causing small I/O accesses at a high rate for packets sized around this cutoff point.
Regardless, this effect is small when observed.

@@

\begin{figure}
	\centering
	\begin{subfigure}{0.45\linewidth}
		\resizebox{1.0\linewidth}{!}{
%			\includegraphics{../plots/build/stress/heat-latency-median}
			\includegraphics{../plots/build/stress/heat-latency-two-9-rel}
%			\includegraphics{../plots/build/stress/heat-latency-two-9-perc}
		}
		\caption{Deviations in \nth{99} percentile cross-traffic RTTs.\label{fig:dataplane-heat}}
	\end{subfigure}
	\hspace{0.05\linewidth}
		\begin{subfigure}{0.45\linewidth}
		\resizebox{1.0\linewidth}{!}{
			\includegraphics{../plots/build/stress/histo-128B-0-3-trim}
		}
		\caption{Distribution of RTTs for \SI{128}{\byte} packets for \numlist{0;3000} RL actions/s.\label{fig:dataplane-example}}
	\end{subfigure}
	\caption{Effects on tail latency of cross-traffic caused by different loads of off-path RL compute. Statistically significant increases in population latency are concentrated on smaller packet sizes, and are typically sub-\SI{78}{\nano\second}.\label{fig:dataplane-coop}}
\end{figure}

\fakepara{Resource requirements}
\Cref{tab:resources} demonstrates how \approachshort{} consumes shared memory as it scales to fit a device's compute resources, compared with a simple P4 forwarding application.
As one program is installed per ME, these results represent the minimum and maximum resource use we can command on a single island (i.e., without replacing P4 workers).
We observe negligible costs on shared bulk EMEM storage ($\sim$\SI{4}{\mebi\byte}), incurred due to hashtables for state-action pairs and rewards.
The most significant costs arise due to policy data (\SI{405}{\kibi\byte} shared IMEM, \SI{90}{\kibi\byte} local CTM, \SI{15}{\kibi\byte} local CLS), which can be halved or quartered using \SIlist{16;8}{\bit} quantisation and remain constant regardless of compute unit usage.
This causes a high upfront cost on per-island resources (CLS/CTM)---\approachshort{} leaves resources for other off-path dataplane applications, but is fairest from 3 cores onwards.

@@

\fakepara{Why in-NIC?}
The above approach relies on co-hosting traffic measurement and analysis alongside OpenFlow-compatible switches at the edge nodes of an AS.
However, packet mirroring, offloading RL computation to a host (potentially over layers of virtualisation), and CPU performance are all sources of additional, consistent state-action latency.
Both traffic statistics collection \emph{and} data-driven learning must be executed on such hosts/network functions.
Unless running these stages in a dedicated pipeline (adding further processing latency), resource contention between these processes will further impact tail latencies.
Naturally, this requires high-performance hardware to be stacked at network egress points, potentially beyond reasonable space, power, or ventilation constraints.
The solution to implement and improve upon this work using \approachshort{} is to place its RL agents on SmartNICs at AS edge nodes---a bump-in-the-wire deployment.
